### 2024-09-12 globals/metadata.tex
enormous rewrite

\newcommand{\makeabstract}{
    \begin{abstract}
This paper studies which functions computed by $\Rel$-weighted automata can be
realised by $\Nat$-weighted automata, under two extra assumptions:
commutativity (the order of letters in the input does not matter) and
polynomial growth (the ouptut of the function is bounded by a polynomial in the
size of the input). This continues the work of Karhumäki, who studied the same
question under the assumption that the computed function was a polynomial,
which is more restrictive. Answering this question starts by exhibiting a
counterexample to the theorem of Karhumäki, proposing a new and valid
characterization in the case of polynomials, that lifts to the more general
setting of polynomial growth and commutativity. Surprisingly, we manage to
leverage this effective characterization to decide whether a function computed
by a commutative $\Nat$-weighted automaton of polynomial growth is star-free, a
notion borrowed from the theory of regular languages that has been the subject
of many investigations in the context of string-to-string functions
during the last decade. 
    \end{abstract}
}

>>>>

\newcommand{\makeabstract}{
    \begin{abstract}
This paper studies which functions computed by $\Rel$-weighted automata can be
realised by $\Nat$-weighted automata, under two extra assumptions:
commutativity (the order of letters in the input does not matter) and
polynomial growth (the ouptut of the function is bounded by a polynomial in the
size of the input). We
leverage this effective characterization to decide whether a function computed
by a commutative $\Nat$-weighted automaton of polynomial growth is star-free, a
notion borrowed from the theory of regular languages that has been the subject
of many investigations in the context of string-to-string functions
during the last decade. 
    \end{abstract}
}

### 2024-09-12 parts/appendix.tex
enormous rewrite

@@

	We will use the fact that given $n_1, \dots, n_p \geq 0$,
	there exists a bijection between
	partitions of $q(\omega n_1, \dots, \omega n_p)$
	into $k+1$ words, and
	triples composed of
	a position  $1 \leq \ell \leq p$,
	a ratio     $0 \leq b \leq n_\ell$,
	a remainder $0 \leq r < \omega \times \card{u_\ell}$,
	and a partition of
	$)u_\ell^\omega)_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{n_p} \alpha_{p}$
	into $k$ words,
	where we have used the notation
	$(v)_{[-r:]}$ to denote the word obtained from $v$
	by keeping only the last $r$ letters, in a sort of Pythonic syntax.
	That is, we look at where the first ``cut" is made.

	Now, given $(\ell, b, r)$ respecting the inequalities
	mentioned above,
	one can define the \kl{polynomial functions}
	$s_{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
	and
	$q_{\ell,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
	so that
	$q(X_1, \dots, X_p) = s_{\ell,r,b}(X_1, \dots, X_{\ell-1})
		q_{\ell, r}(X_{\ell} - b, \dots, X_p)$
	under the assumption that $X_\ell \geq b$:
	\begin{equation*}
		q_{\ell,r}(X_{1}, \dots, X_{p - \ell + 1})
		\defined
		u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times X_1}
		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
		\quad .
	\end{equation*}
	Similarly, let us define
	$s_{\ell, r, b}(X_1, \dots, X_{\ell - 1})$
	as
	$\alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1}
		u_\ell^{\omega b} (u_{\ell}^\omega)_{[1:r]}$,
	Remark that $q_{\ell,r}$ has all of its indeterminates
	multiplied by $\omega$, so that the induction hypothesis
	can be used.

	The key remark is that $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell - 1}))$ is a
	constant function from $\Nat^{\ell-1}$ to $M$, and that its value
	only depends on whether $b = 0$ or $b > 0$.
	This holds with
	the exact same proof as the base case of this induction. Let us call
	$m_{\ell,r,1}$ and $m_{\ell,r,0}$ the two elements of the monoid that are
	obtained by $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell-1}))$ respectively when
	$b = 0$ and when $b > 0$. Finally, let us define $\pi_{\ell,r,0} \colon M^k
		\to \Nat$ via $\pi_{\ell,r,0}(m_2, \dots, m_{k+1}) \defined
		\pi(m_{\ell,r,0}, m_1, \dots, m_{k+1})$.

	With all of this preliminary work done, we are now ready to conclude by
	simply unrolling the definition of
	a function
	$f$ defined using a \kl{production function}
	$\pi \colon M^{k+1} \to \Nat$.
	In the following equations, we omit the precise
	range for the sum over $(\ell, r, b)$ to save space.
	Similarly, we will use
	$q_{\ell,r}(X_1, \dots, X_{p-\ell+1})
		\defined q_{\ell,r}(X_1+b, X_2, \dots, X_{p - \ell + 1})$
	at the end of the computation.
	\begin{align*}
		 & f(q(n_1, \dots, n_p))           \\
		 & \eqdef
		\sum_{v_1 \dots v_{k+1} = q(n_1, \dots, n_p)}
		\pi(\mu(v_1), \dots, \mu(v_{k+1})) \\
		 & = \sum_{(\ell, b, r)}
		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
		\pi(\mu(s_{\ell,r,b}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & =
		\sum_{(\ell, 0, r)}
		\sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
		\pi(m_{\ell,r,0}, \mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & +
		\sum_{(\ell, b > 0, r)}
		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
		\pi(m_{\ell,r,1}, \mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & =
		\sum_{(\ell, 0, r)}
		\sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
		\pi_{\ell,r,0}(\mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & +
		\sum_{(\ell, b > 0, r)}
		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell-b, \dots, n_p)}
		\pi_{\ell,r,1}(\mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & =
		\sum_{1 \leq \ell \leq p}
		\sum_{0 \leq r < \omega \times \card{u_\ell}}
		(\pi_{\ell,r,0})^\dagger (q_{\ell, 0, r}(n_\ell-b, \dots, n_p))
		\\
		 & +
		\sum_{1 \leq \ell \leq p}
		\sum_{0 \leq r < \omega \times \card{u_\ell}}
		\sum_{0 < b < n_\ell}
		(\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
	\end{align*}
	Now, using the induction hypothesis,
	we have \kl{natural binomial functions} $F_{\ell,r,0}$ and $F_{\ell,r,1}$
	satisfying
	\begin{equation*}
		(\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
		= F_{\ell,r,1}(\omega (n_\ell - b), \dots, \omega n_p)
		\quad
		\forall n_\ell - b, \dots, n_p \geq 1
		\quad .
	\end{equation*}
	And similarly
	\begin{equation*}
		(\pi_{\ell,r,0})^{\dagger}(q_{\ell,r}(n_1, \dots, n_{\ell-1})
		= F_{\ell,r,0}(\omega n_1, \dots, \omega n_{\ell-1})
		\quad
		\forall n_1, \dots, n_{\ell - 1} \geq 1
		\quad .
	\end{equation*}

@@

\begin{proofof}[natural-binomial-polynomial-positive-bad:ex]
    $Q$ belongs to $\Rel[\vec{X}]$, but not to $\CorrectPoly$ because
    $Q(0,Y,Z)$ has a negative maximal monomial,
    hence it cannot be \kl{represented} by a \kl{star-free $\Nat$-polyregular function}
    thanks to \cref{corrected-version:thm}.
\end{proofof}

\section{Proofs of section \ref{beyond-polynomials:sec}}

\begin{proofof}[decompose-polynomial:lem]
    Let us first write $f \in \ZPoly$ as a linear combination 
    $f = f_+ - f_-$ where both $f_+$ and $f_-$ are in $\NPoly$.
    We 
    compute $\omega_+$ (resp. $\omega_-$)
    by
    applying
    \cref{n-poly-combinatorics:lem} 
    to $f_+$ (resp. to $f_-$).
    Finally, let $\omega$ be the least common multiple
    of $\omega_+$ and $\omega_-$, i.e. $\omega \defined \ppcm(\omega_+, \omega_-)$.

    Using this $\omega$,
    for any choice of
    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
    there exists two \kl{natural binomial functions} $F_r^+$ and $F_r^-$
    associated to $f_+$ and $f_-$,
    such that $x_1, \dots, x_p \geq 1$:
    \begin{equation*}
        f\left(
        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right) 
        = F_r^+(x_1, \dots, x_p) - F_r^-(x_1, \dots, x_p) \quad .
    \end{equation*}
    Now, for large enough values of $x_1, \dots, x_p$,
    this means that 
    $f\left(
        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right)$
    coincides with an \kl{integer binomial polynomial},
    and we can therefore conclude selecting different values of $r$ and
    setting some subsets of variables to be assigned to the constant $0$.
\end{proofof}

>>>>

\begin{proofof}[n-poly-n-poly:example]
    The polynomials of $\Nat[\vec{X}]$
    are obtained from basic functions (constant in $\Nat$,
    $X_i$ for some $1 \leq i \leq n$)
    by products and sums (\cref{stability-polyregular:lemma}). Because the basic functions are
    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions (see
    \cref{size-of-word-nsf:ex}),
    that are closed under these operations, we conclude.
\end{proofof}

@@

	We will use the fact that given $n_1, \dots, n_p \geq 0$,
	there exists a bijection between
	partitions of $q(\omega n_1, \dots, \omega n_p)$
	into $k+1$ words, and
	triples composed of
	a position  $1 \leq \ell \leq p$,
	a ratio     $0 \leq b \leq n_\ell$,
	a remainder $0 \leq r < \omega \times \card{u_\ell}$,
	and a partition of
	$(u_\ell^\omega)_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{n_p} \alpha_{p}$
	into $k$ words,
	where we have used the notation
	$(v)_{[-r:]}$ to denote the word obtained from $v$
	by keeping only the last $r$ letters, in a sort of Pythonic syntax.
	That is, we look at where the first ``cut" is made.

	Now, given $(\ell, b, r)$ respecting the inequalities
	mentioned above,
	one can define the \kl{polynomial functions}
	$\CutPref{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
	and
	$\CutSuff{\ell,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
	so that
	$q(X_1, \dots, X_p) = \CutPref{\ell,r,b}(X_1, \dots, X_{\ell-1})
		q_{\ell, r}(X_{\ell} - b, \dots, X_p)$
	under the assumption that $X_\ell \geq b$:
	\begin{align*}
        \CutPref{\ell,r,b}(X_1, \dots, X_{\ell - 1})
        &\defined
        \alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1}
        u_\ell^{\omega b} (u_{\ell}^\omega)_{[1:r]}
        \quad .
        \\
		\CutSuff{\ell,r}(X_{1}, \dots, X_{p - \ell + 1})
        &\defined
        (u_\ell^\omega)_{[-r:]} u_\ell^{\omega X_1}
		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
		\quad .
	\end{align*}
	Remark that $\CutSuff{\ell,r}$ has all of its indeterminates
	multiplied by $\omega$, so that the induction hypothesis
	can be used.

	The key remark is that $\mu(\CutPref{\ell,r,b}(X_1, \dots, X_{\ell - 1}))$ is a
	constant function from $\Nat^{\ell-1}$ to $M$, and that its value
	only depends on whether $b = 0$ or $b > 0$.
	This holds with
	the exact same proof as the base case of this induction. Let us call
	$m_{\ell,r,1}$ and $m_{\ell,r,0}$ the two elements of the monoid that are
	obtained by $\mu(\CutPref{\ell,r,b}(X_1, \dots, X_{\ell-1}))$ respectively when
	$b = 0$ and when $b > 0$. Finally, let us define $\pi_{\ell,r,0} \colon M^k
		\to \Nat$ via $\pi_{\ell,r,0}(m_2, \dots, m_{k+1}) \defined
		\pi(m_{\ell,r,0}, m_1, \dots, m_{k+1})$.

	With all of this preliminary work done, we are now ready to conclude by
	simply unrolling the definition of
	a function
	$f$ defined using a \kl{production function}
	$\pi \colon M^{k+1} \to \Nat$.
	In the following equations, we omit the precise
	range for the sum over $(\ell, r, b)$ to save space.
	\begin{align*}
		 & f(q(n_1, \dots, n_p))           \\
		 & \eqdef
		\sum_{v_1 \dots v_{k+1} = q(n_1, \dots, n_p)}
		\pi(\mu(v_1), \dots, \mu(v_{k+1})) \\
		 & = \sum_{(\ell, b, r)}
		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,r}(n_\ell - b, \dots, n_p)}
		\pi(\mu(\CutPref{\ell,r,b}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & =
		\sum_{(\ell, 0, r)}
		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,b,0}(n_\ell - b, \dots, n_p)}
		\pi(m_{\ell,r,0}, \mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & +
		\sum_{(\ell, b > 0, r)}
		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,r}(n_\ell - b, \dots, n_p)}
		\pi(m_{\ell,r,1}, \mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & =
		\sum_{(\ell, 0, r)}
		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,b,0}(n_\ell - b, \dots, n_p)}
		\pi_{\ell,r,0}(\mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & +
		\sum_{(\ell, b > 0, r)}
		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,r}(n_\ell-b, \dots, n_p)}
		\pi_{\ell,r,1}(\mu(v_2), \dots, \mu(v_{k+1}))
		\\
		 & =
		\sum_{1 \leq \ell \leq p}
		\sum_{0 \leq r < \omega \times \card{u_\ell}}
		(\pi_{\ell,r,0})^\dagger (\CutSuff{\ell, 0, r}(n_\ell-b, \dots, n_p))
		\\
		 & +
		\sum_{1 \leq \ell \leq p}
		\sum_{0 \leq r < \omega \times \card{u_\ell}}
		\sum_{0 < b < n_\ell}
		(\pi_{\ell,r,1})^{\dagger}(\CutSuff{\ell,r}(n_\ell - b, \dots, n_p))
	\end{align*}
	Now, using the induction hypothesis,
	we have \kl{natural binomial functions} $F_{\ell,r,0}$ and $F_{\ell,r,1}$
	satisfying
	\begin{equation*}
		(\pi_{\ell,r,1})^{\dagger}(\CutSuff{\ell,r}(n_\ell - b, \dots, n_p))
		= F_{\ell,r,1}(\omega (n_\ell - b), \dots, \omega n_p)
		\quad
		\forall n_\ell - b, \dots, n_p \geq 1
		\quad .
	\end{equation*}
	And similarly
	\begin{equation*}
		(\pi_{\ell,r,0})^{\dagger}(\CutSuff{\ell,r}(n_1, \dots, n_{\ell-1})
		= F_{\ell,r,0}(\omega n_1, \dots, \omega n_{\ell-1})
		\quad
		\forall n_1, \dots, n_{\ell - 1} \geq 1
		\quad .
	\end{equation*}

@@

\begin{proofof}[polyrec-integer-strong:lem]
    Because $n P \in \CorrectPoly$,
    it is represented by a \kl{star-free $\Nat$-polyregular function} $f$ thanks to
    \cref{corrected-version:thm}. Let us now use \cref{n-sf-combinatorics:lem}
    to obtain a number $s \geq 1$ (that only depends on $P$) such
    that for all \kl{polynomial function} $q \colon \Nat^k \to \Sigma^*$, $g
    \circ q(X_1 + s, \dots, X_k + s)$ is a \kl{natural binomial function} $F$
    over $\Nat^k$. Let us write the
    \kl{natural binomial function} $F$ as follows $F(X_1, \dots, X_k) =
    \sum_{\card{\alpha} \leq d_2} e_{\alpha} \prod_{i = 1}^k \binom{X -
    p_{\alpha,i}}{\alpha(i)}$, with $e_{\alpha} \in \Nat$ for all
    $\card{\alpha} \leq d_2$.
    Now, because of the equality 
    $\binom{X - p - 1}{k} + \binom{X - p - 1}{k - 1} = \binom{X - p}{k}$,
    we conclude that we can find a $K \in \Nat$ and coefficients
    $f_{\alpha} \in \Nat$ such that 
    $F(X_1, \dots, X_k) =
    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \binom{X_i - K}{\alpha(i)}$.
    Now, remark that $P$ is an \kl{integer-valued} polynomial of
    degree $d$, and therefore one can write $P = \sum_{\card{\alpha} \leq d}
    c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$ with $c_{\alpha}
    \in \Rel$ for all $\card{\alpha} \leq d$
    (\cref{integer-binomial-polynomial:lem}). 
    As a consequence,
    $\lambda P = \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K - s}{\alpha(i)}$,
    and 
    by identifying coefficients in the previous two expressions, we obtain
    $c_{\alpha} = \lambda f_{\alpha}$ for all $\card{\alpha} \leq d$.
    Finally,
    $\lambda P(x_1 + s, \dots, x_k + s) = F(x_1, \dots, x_k)$ for all $\vec{x} \in \Nat^k$,
    and therefore
    $\lambda P (X_1 + s, \dots, X_k + s) =
    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$.
    In particular,
    we get
    \begin{equation*}
        \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)} =
        \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
    \end{equation*}
    And we conclude by term identification
    by term identification
    that $d = d_2$, and for all $\card{\alpha} \leq d$, $\lambda c_{\alpha} = f_{\alpha}$.
    As a consequence, $c_{\alpha} \in \Nat$ for all $\card{\alpha} \leq d$,
    and $P$
    is a \kl{natural binomial polynomial}.
    Remark that this reasoning remains true whenever we partially apply $P$ to some values
    in $\Nat$, as it amounts to 
    partially applying the function $f$. We have proven that 
    $P$ is a \kl{strongly natural binomial polynomial}.
\end{proofof}

### 2024-09-12 parts/conclusion.tex
enormous rewrite

\begin{lemma}
    \proofref{pre-compose-growth-commut:lemma}
    \label{pre-compose-growth-commut:lemma}
    Let $f \in \ZRat$, and $k \in \Nat$. Then,
    $f \in \ZPoly[k]$ if and only if 
    for every \kl{commutative} \kl{star-free polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[k\times l]$.
\end{lemma}

\begin{lemma}
    \proofref{pre-compose-sf-commut:lemma}
    \label{pre-compose-sf-commut:lemma}
    Let $f \in \ZPoly$. Then, $f \in \ZSF$,
    if and only if for every \kl{commutative} \kl{star-free polyregular function} $h$,
            $(f \circ h) \in \ZSF$.
\end{lemma}

Remark that if those two lemmas were to hold for \kl{$\Nat$-polyregular
functions}, then the decidability of $\NPoly$ inside $\ZPoly$ would follow
immediately, and $\NSF$ inside $\NPoly$ would also immediately follow. On the
one hand, one can guess a candidate function and check for equivalence, on the
other hand, one can guess a \kl{commutative} \kl{star-free polyregular
function} and check membership (which is decidable thanks to this paper).

>>>>

\begin{lemma}[restate=pre-compose-growth-commut:lemma,label=pre-compose-growth-commut:lemma]
    \proofref{pre-compose-growth-commut:lemma}
    Let $f \in \ZRat$, and $k \in \Nat$. Then,
    $f \in \ZPoly[k]$ if and only if 
    for every \kl{commutative} \kl{star-free polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[k\times l]$.
\end{lemma}

\begin{lemma}[restate=pre-compose-sf-commut:lemma,label=pre-compose-sf-commut:lemma]
    \proofref{pre-compose-sf-commut:lemma}
    Let $f \in \ZPoly$. Then, $f \in \ZSF$,
    if and only if for every \kl{commutative} \kl{star-free polyregular function} $h$,
            $(f \circ h) \in \ZSF$.
\end{lemma}

Remark that if \cref{pre-compose-sf-commut:lemma} were to hold for
\kl{$\Nat$-polyregular functions}, then the decidability of $\NPoly$ inside
$\ZPoly$, and the decidability of $\NSF$ inside $\NPoly$ would immediately
follow. On the one hand, one can guess a candidate function and check for
equivalence, on the other hand, one can guess a \kl{commutative} \kl{star-free
polyregular function} and check membership (which is decidable thanks to this
paper). This is restated in our concluding conjecture.

\begin{conjecture}
    \label{npoly-zpoly:conjecture}
    Let $k \in \Nat$.
    A function $f \in \ZPoly[k]$
    belongs to $\NPoly[k]$ if and only if
    it is \kl{$(k,\Nat)$-combinatorial}.
\end{conjecture}
